\documentclass[a4paper,11pt]{article}
\usepackage[T1]{fontenc}
\usepackage[u  tf8]{inputenc}
\usepackage{mathtools}
\usepackage{mathrsfs}
\usepackage{bm}
\usepackage{amsfonts}
\usepackage[dvipsnames]{xcolor}
% \usepackage{cleveref}
\usepackage[normalem]{ulem}
\usepackage{graphicx}
\usepackage{fullpage}
\usepackage[margin=1in]{geometry}
\usepackage{natbib}
\usepackage{amsmath}
%\usepackage{float}
%\usepackage{subcaption}
%\usepackage{multirow} 
\usepackage[colorlinks = true,urlcolor=blue]{hyperref}
%\usepackage{bibunits}
%\usepackage{csvsimple}
%\usepackage[superscript,biblabel]{cite}
\usepackage{verbatim}
\usepackage{graphicx}
\graphicspath{ {images/} }
\usepackage{subfig}
\usepackage{subcaption}

\newcommand{\jb}[1]{{\color{blue} (#1)} }
\begin{document}

\title{Title here}
\author{Kyelee Ruth Fitts, Jeremy Berg, Guy Amster, Guy Sella}
\maketitle


\subsection*{Abstract}

\subsection*{Introduction}
It has long been known that diversity on Earth is in large part due to the
adaptation of organisms to varying environments. We now know that
adaptation has a large genetic basis, and that genetic variation is
the key to biological diversity. Thus, understanding how and why such
genetic variation occurs is the ultimate goal of evolutionary
biology. Quantifying genetic adaptation is the goal of population
geneticists, and to this end finding methods to detect signals of
selection within the genome is of particular interest. Change in
variation over time can be attributed to many causes including genetic
drift or population migration, so detecting signals of selection in
particular is no trivial task. 

Wright first introduced the parameter $F_{st}$ as a measure comparing
the variation of an allele in a subpopulation relative to the entire population
\cite{Fst}. Lewontin and Krakauer, using this parameter developed a
novel statistical test based on the fact
that under no selection, the variation of specific alleles over all
subpopulations will not be different from that of the population
$F_{st}$ \cite{firstseltest}. These conclusions were extended by Spitze, who coined the
parameter $Q_{st}$ as a measure of how the phenotypic value of an
allele in a subpopulation compares to that of the entire
population \cite{Qst}. These early tests for selection utilized the ratio
$\frac{Q_{st}}{F_{st}}}$ as a test statistic for detecting signals of
selection, where $F_{st} = Q_{st}$ is the null model, where no
selection occurs. 

Until recently, the study of detecting selection has been limited to
large-effect alleles, when in fact many phenotypic traits of interest
are affected by many different loci across the genome. Understanding
how to find signals of selection for such polygenic traits has been
the subject of much study in recent years. Increased compuational
capabilities have given us to tools to be able to evaluate the effects
of many thousands of loci on a trait, allowing us to evaluate
selection in highly polygenic traits, such as height. One of the most
important tools developed are Genome-Wide Association Studies (GWAS)
which determine which loci across the genome have significant effects
on a certain polygenic trait. 

Berg introduces a comprehensive test statistic, called $Q_x$, to test
for selection \cite{gwasintro}. This statistic depends on $F_{st}$ and
a generalized analogy to $Q_{st}$, expressed in terms of $\alpha$, or
the effect size of each significant locus as determined by GWAS, and
the allele frequencies $p$ over loci $l$ and $l'$, summed over
populations $m$. $V_a$ is the additive genetic variance of the entire population.

\begin{equation} \label{eqn:Qxraw}
  Q_X = \frac{1}{V_A F_{ST}} \sum_{m=1}^M \sum_{\ell=1}^L \sum_{\ell\prime=1}^L \alpha_{\ell} \alpha_{\ell^{\prime}}\left(p_{m\ell} - \overline{p}_\ell \right)\left(p_{m \ell\prime} - \overline{p}_{\ell\prime}\right)
\end{equation}

The distribution of this statistic is expected to be chi-squared. We are particularly interested in the average effect of this
statistic-- not only does GWAS use it to determine which alleles are deemed
significant to a trait, but it is also the weighting factor for loci in the
expression for $Q_x$. The average effect size depends on the allele
frequency, the homozygous effect size (A), or effect size of the most frequent homozygote, and the
dominance deviation (D), which is the difference in effect size for the heterozygote
deviating from $\frac{1}{2}$ of the homozygous effect size.

\begin{equation}  
  \alpha_\ell = \frac{1}{2} A_\ell + D_\ell\left(1-2p_{1\ell}\right).
  \label{avgeff}
\end{equation}

The use of $\alpha$ in this test statistic becomes problematic in the
presence of directional dominance, when alleles with a postive effect
size on the trait are systematically dominant and alleles with a
negative effect size on the trait are systematically negative. In
terms of \eqref{avgeff}, this means that the signs of A and D are the
same for a large number of loci.

In the presence of directional dominance, we hypothesized a bias in
tests for polygenic adaptation that depend on $\alpha$. Alleles which
are systematically dominant and which have recently increased in
frequency (large p, positive D) will tend to have smaller effect sizes, while
alleles which are recessive which have recently decreased in frequency
(small p, negative D) will tend to have larger effect sizes. In the
latter case, the test for polygenic selection advanced by Berg
\eqref{eqn:Qxraw} will be apt to make false positive judgements for
selection, because the statistic will be calculated over alleles which
do not actually increase the effect size.

Height is one polygenic trait with many well-defined signficant
alleles via GWAS \cite{heightselection}. It has also been shown to exhibit
directional dominance \cite{heightdirectdom}. We aimed to quantify the
hypothesized bias, first with simulated populations, then with height genotype data from the UK Biobank.



\subsection*{Theory/Methods}

\subsection*{Results}

\subsection*{Discussion}


\bibliography{works_cited}
\bibliographystyle{plain}






\end{document}
